---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
```

# `wgsLR`: Shotgun sequencing for human identification: Dynamic SNP marker sets and likelihood ratio calculations accounting for errors

Please refer to the online documentation at <https://mikldk.github.io/wgsLR>, including the vignettes.

## Scientific publication

The research associated with this software is described in 

> Andersen, M. M., Kampmann, M.-L., Jepsen, A. H., Morling, N., Eriksen, P. S., Børsting, C., & Andersen, J. D. (2025). 
> *Shotgun DNA sequencing for human identification: Dynamic SNP selection and likelihood ratio calculations accounting for errors.* 
> Forensic Science International: Genetics, 74, 103146. [doi:10.1016/j.fsigen.2024.103146](https://doi.org/10.1016/j.fsigen.2024.103146).


## Installation

To install from Github with vignettes run this command from 
within `R` (please install `remotes` first if not already installed):

```
# install.packages('remotes')
remotes::install_github("mikldk/wgsLR", build_vignettes = TRUE)
```

You can also install the package without vignettes if needed as follows:

```
remotes::install_github("mikldk/wgsLR")
```


## A few small examples

```{r, message=FALSE, warning=FALSE}
library(wgsLR)
```


### Estimating the genotype error probability, $w$

```{r}
cases <- wgsLR::sample_data_Hp_w(n = 1000, w = 0.1, p = c(0.25, 0.25, 0.5))
tab <- table(cases$xT, cases$xR)
tab
w_mle <- wgsLR::estimate_w(tab)
w_mle
```


#### Cautionary note: not just standard VCF files

It is necessary to obtain sequencing results for all bases in the
selected segments (including read depth and genotype quality). 
Thus, it is **not** sufficient to just use information from
"confirmed"/high probability variants from the reference genome (variants identified in standard vcf-file format), as this can introduce bias in the results. 
**Information from all bases in the chosen genomic areas of interest is needed**. 
One way to achieve this by using [GATK HaplotypeCaller](https://gatk.broadinstitute.org/hc/en-us/articles/9570334998171-HaplotypeCaller) 
with the additional argument `--emit-ref-confidence BP_RESOLUTION` for the 
genomic areas of interest (using `-L areas.interval_list`).



### Calculating likelihood ratios ($LR$'s)

Same genotyping error probability for both trace sample ($x_t =$ `xT`) and reference sample ($x_r =$ `xR`):

```{r}
tab <- wgsLR::d_LR_w[, c("xT", "xR")]
tab$LR <- paste0("$", wgsLR::d_LR_w$expr_tex, "$")
tab |> 
  kbl(format = "markdown")
```

Sample-specific genotyping error probabilities $w_t$ and $w_r$ for trace sample ($x_t =$ `xT`) and reference sample ($x_r =$ `xR`), repectively:

```{r}
tab <- wgsLR::d_LR_wTwR[, c("xT", "xR")]
tab$LR <- paste0("$", wgsLR::d_LR_wTwR$expr_tex, "$")
tab |> 
  kbl(format = "markdown")
```

#### Example

Assume that a trace sample had four loci with genotypes (0/1 = 1, 0/0 = 0, 1/1 = 2, 1/1 = 2).
A person of interest is then typed for the same four loci and 
has genotypes (0/0 = 0, 0/0 = 0, 1/1 = 2, 1/1 = 2), i.e. a mismatch on the first locus.

For simplicity, assume that the genotype probabilites are 
P(0/0 = 0) = 0.25, 
P(0/1 = 1/0 = 1) = 0.25, and 
P(1/1 = 2) = 0.5.

If no errors are possible, then $w=0$ and

```{r}
wgsLR::calc_LRs_w(xT = c(1, 0, 2, 2),
                  xR = c(0, 0, 2, 2), 
                  w = 0, 
                  p = c(0.25, 0.25, 0.5))
```

and the product is 0 due to the mismatch at the first locus.

If instead we acknowledge that errors are possible, then for $w = 0.001$ we obtain that

```{r}
LR_contribs <- wgsLR::calc_LRs_w(xT = c(1, 0, 2, 2), 
                                 xR = c(0, 0, 2, 2), 
                                 w = 0.001, 
                                 p = c(0.25, 0.25, 0.5))
LR_contribs
prod(LR_contribs)
```

We can also consider the $LR$s for a range for plausible values of $w$:

```{r}
ws <- c(1e-6, 1e-3, 1e-2, 1e-1)
LRs <- sapply(ws, \(w) wgsLR::calc_LRs_w(xT = c(0, 0, 2, 2), 
                                         xR = c(1, 0, 2, 2), 
                                         w = w, 
                                         p = c(0.25, 0.25, 0.5)) |> 
                prod())
data.frame(log10w = log10(ws), w = ws, 
           LR = LRs, WoElog10LR = log10(LRs))
```


## Different error rates

Assume that the trace donor profile has $w_T = 10^{-4}$ and 
the suspect reference profile has $w_R = 10^{-8}$. Then 
the $LR$ is:

```{r}
LR_contribs <- wgsLR::calc_LRs_wTwR(xT = c(1, 0, 2, 2), 
                                    xR = c(0, 0, 2, 2), 
                                    wT = 1e-4, 
                                    wR = 1e-8,
                                    p = c(0.25, 0.25, 0.5))
prod(LR_contribs)
sum(log10(LR_contribs))
```


## Unknown trac sample error rate, $w_T$


Assume that the trace donor profile has unknown $w_T$ but 
can be assumed to follow a beta prior on $(0, \frac 1 2)$ with 
mean value $\mu = 10^{-4}$ and variance $\sigma^2 = \mu^2 / 2$:

```{r}
mu <- 1e-4
sigmasq <- mu^2 / 2
shpT <- wgsLR::get_beta_parameters(mu = mu, sigmasq = sigmasq, a = 0, b = 0.5)
# 90% probability:
q <- wgsLR::qbeta05(c(0.05, 0.95), shape1 = shpT[1], shape2 = shpT[2])
q
```

Further, assume that the suspect reference profile has $w_R = 10^{-8}$. 
Then the $WoE = \log_{10}(LR)$ is:

```{r}
wgsLR::calc_WoE_wTwR_integrate_wT_mc(
  xT = c(1, 0, 2, 2), 
  xR = c(0, 0, 2, 2), 
  shape1T_Hp = shpT[1], shape2T_Hp = shpT[2],
  shape1T_Ha = shpT[1], shape2T_Ha = shpT[2],
  wR = 1e-8,
  p = c(0.25, 0.25, 0.5))
```
