---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(wgsLR)
```

# Estimating the genotype error probability, $w$

A simple example:

```{r}
cases <- wgsLR::sample_data_Hp_w(n = 1000, w = 0.1, p = c(0.25, 0.25, 0.5))
tab <- table(cases$X_D, cases$X_S)
tab
```

## Maximum likelihood estimation

```{r}
w_mle <- wgsLR::estimate_w(tab)
w_mle
w_mle_se <- wgsLR::estimate_w_se(tab, w_mle)
w_mle_se
```

The last lines estimate the standard error that will be used below to construct an approximate confidence interval.

*Please be aware of the "Cautionary note" in the README (about not just using standard VCF-files).*

## Bayesian inference

The package contains a simple Metropolis-Hastings sampler (also see the "Using Stan for Bayesian inference" vignette):

```{r}
w_b <- wgsLR::estimate_w_bayesian(tab) |> posterior_samples()
w_b_q <- quantile(w_b, c(0.025, 0.975))
```

```{r}
hist(w_b, main = NULL, xlab = expression(hat(w)))
abline(v = mean(w_b), col = "deeppink3", lwd = 3)
abline(v = w_b_q[1], col = "deeppink3", lty = 2)
abline(v = w_b_q[2], col = "deeppink3", lty = 2)
abline(v = w_mle, col = "yellow3", lwd = 3)
abline(v = w_mle - 2*w_mle_se, col = "yellow3", lty = 2)
abline(v = w_mle + 2*w_mle_se, col = "yellow3", lty = 2)
legend("topright", col = c("deeppink3", "yellow3"), legend = c("Bayesian", "MLE"), lty = 1, lwd = 2)
```

The dashed lines indicate approximate confidence intervals (maximum likelihood estimation) and 
credible interval (Bayesian inference).

*Please be aware of the "Cautionary note" in the README (about not just using standard VCF-files).*

# Calculating likelihood ratios ($LR$'s)


No errors and matching genotypes:

```{r}
LR_contribs <- wgsLR::calc_LRs_w(xs = c(0, 0), 
                                 xd = c(0, 0), 
                                 w = 0, 
                                 p = c(0.25, 0.25, 0.5))
prod(LR_contribs)
```

No errors and non-matching genotypes:

```{r}
LR_contribs <- wgsLR::calc_LRs_w(xs = c(0, 0), 
                                 xd = c(1, 0), 
                                 w = 0, 
                                 p = c(0.25, 0.25, 0.5))
prod(LR_contribs)
```


Errors possible and matching genotypes:

```{r}
LR_contribs <- wgsLR::calc_LRs_w(xs = c(0, 0), 
                                 xd = c(0, 0), 
                                 w = 0.001, 
                                 p = c(0.25, 0.25, 0.5))
prod(LR_contribs)
```

Errors possible and non-matching genotypes:

```{r}
LR_contribs <- wgsLR::calc_LRs_w(xs = c(0, 0), 
                                 xd = c(1, 0), 
                                 w = 0.001, 
                                 p = c(0.25, 0.25, 0.5))
prod(LR_contribs)
```

## Different error rates

Assume that the trace donor profile has $w_D = 10^{-4}$ and 
the suspect reference profile has $w_S = 10^{-8}$. Then 
the $LR$ is:

```{r}
LR_contribs <- wgsLR::calc_LRs_wDwS(xs = c(0, 0), 
                                    xd = c(1, 0), 
                                    wD = 1e-4, 
                                    wS = 1e-8,
                                    p = c(0.25, 0.25, 0.5))
prod(LR_contribs)
```

# Verifying formulas

The usual $LR$ values, $\frac{1}{p_{Z^S}}$ for matches and $0$ for non-matches, 
are obtained when setting $w = 0$.
For example with both packages `wgsLR` and `caracas` loaded we can do the following:

```{r, eval = require("caracas", quietly = TRUE), warning=FALSE}
library(caracas)
LR <- d_prob_LR_w$expr
LRw0 <- sapply(LR, \(x) as_sym(x) |> subs("w", 0) |> as_character())
with(d_prob_LR_w, cbind(XD_MA, XS_MA, LRw0))
```


And with sample specific error rates, $w_D$ and $w_S$:

```{r, eval = require("caracas", quietly = TRUE), warning=FALSE}
library(caracas)
LR <- d_prob_LR_wDwS$expr
LRw0 <- sapply(LR, \(x) as_sym(x) |> 
                 subs("wD", 0) |> 
                 subs("wS", 0) |> as_character())
with(d_prob_LR_wDwS, cbind(XD_MA, XS_MA, LRw0))
```
