---
title: "Posterior likelihood ratio distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posterior likelihood ratio distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(wgsLR)
```

The following is based on Section 3 in @MeesterSlooten2021.

Assuming everything except trace sample error rate, $w_t$, is known (i.e. reference sample error rate $w_r$ and genotype probabilities $p_0$, $p_1$, and $p_2$), then

\begin{align}
    LR(E) &= \frac{\int P(E \mid w_t, H_p) f(w_t \mid H_p) d w_t}{\int P(E \mid w_t, H_a) f(w_t \mid H_a) d w_t} .
\end{align}

Assuming in the rest of this document that the prior of $w_t$ under $H_p$ and $H_a$ coincide such that $f(w_t \mid H_p) = f(w_t \mid H_a)$ then

\begin{equation}
    LR(E) = \int LR_{w_t}(E) \, f(w_t \mid E, H_a) \, d w_t ,
\end{equation}
where
\begin{align}
    LR_{w_t}(E) &= \frac{P(E \mid w_t, H_p)}{P(E \mid w_t, H_a)} .
\end{align}

Thus, 
\begin{equation}
    LR(E) = E_{w_t \mid E, H_a} \left [ LR_{w_t}(E) \right ] .
\end{equation}

So if we can sample from $w_t \mid E, H_a$ and calculate the $LR$'s as if $w_t$ was known, we get 
the posterior likelihood ratio distribution under $H_a$. 


There are some interesting points in @MeesterSlooten2021 that are quoted below. 
Quotation marks were not used as $\theta$ was replaced by $w_t$ and $H_2$ by $H_a$ and points about the posterior likelihood ratio distribution under $H_p$ was removed:

* Thus, sampling from the posterior likelihood ratio distribution under $H_a$ tells us
what likelihood ratios we expect to obtain if we can eliminate our uncertainty about
$w_t$, but we do not expect the likelihood ratio to change in either direction.

* When we repeatedly sample $LR_{w_t}(E)$ using $f(w_t \mid E, H_a)$,
we can, by disregarding the, say, 5% most extreme outcomes, construct an interval.
This interval tells us something about the values of future likelihood ratios that
we would obtain, were we to remove all existing uncertainty about the model
parameters $w_t$.

### Example

```{r, cache=TRUE}
set.seed(2026)
m <- 100
q <- runif(m, min = 0.2, max = 0.8)
p <- lapply(q, \(q0) c(q0^2, 2*q0*(1-q0), (1-q0)^2))
wR <- 1/100
cases <- sample_data_Ha_wTwR(n = m, wT = 1/10, wR = wR, p = p)
xT <- cases$xT[, 1L]
xR <- cases$xR[, 1L]
table(xT, xR)
```

We can calculate $LR$'s, or rather $WoE = \log_{10}(LR)$:

```{r}
WoE_full_knowledge <- calc_LRs_wTwR(
  xT = xT, 
  xR = xR, 
  wR = wR, 
  wT = 1/10,
  p = p
) |> 
  log10() |> 
  sum()
WoE_full_knowledge
WoE_marginalisation <- calc_WoE_wTwR_integrate_wT_mc(
  xT = xT, 
  xR = xR, 
  wR = wR, 
  p = p,
  shape1T_Hp = 1, 
  shape2T_Hp = 1,
  shape1T_Ha = 1, 
  shape2T_Ha = 1
)
WoE_marginalisation
```


```{r}
wTs <- seq(1/100, 1/2, length.out = 50)
WoEs <- lapply(wTs, \(wT) {
  calc_LRs_wTwR(
    xT = xT, 
    xR = xR, 
    wR = wR, 
    wT = wT,
    p = p
  ) |> 
    log10() |> 
    sum()
}) |> unlist()
plot(wTs, WoEs, type = "b")
abline(h = 0, lty = 2)
abline(v = 1/10, lty = 2)
```


```{r, eval=requireNamespace("cmdstanr", quietly = TRUE), cache=TRUE}
m <- wT_posterior_Ha_stan(xT = xT, xR = xR, wR = wR, p = p,
                          prior_a = 1, 
                          prior_b = 1,
                          chains = 5,
                          parallel_chains = 5,
                          iter_warmup = 100,
                          iter_sampling = 200)
wT_Ha_E <- m$draws("wT") |> c()
hist(wT_Ha_E)
abline(v = mean(wT_Ha_E))
mean(wT_Ha_E)
posterior_WoEs <- lapply(wT_Ha_E, \(wT) {
  calc_LRs_wTwR(
    xT = xT, 
    xR = xR, 
    wR = wR, 
    wT = wT,
    p = p
  ) |> 
    log10() |> 
    sum()
}) |> unlist()
hist(posterior_WoEs)
quantile(posterior_WoEs, c(0.025, 0.975))
```

## References
